---
title: "Great Graphs and Tables in ggplot and kable"
subtitle: "MCRS Quantitative Methods""
author: "Amy Henry"
date: "November 5, 2024"
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
install.packages("ggtext")

library(tidyverse)
library(ggtext)
```

So we all know by now how to make INFORMATIVE graphs using ggplot.

What if we want to make *beautiful* ones?

First let's make some basic plots to play with:

This is a plot of height of pine trees measured repeatedly at different ages.

```{r Basic Plot, warning=FALSE}
Loblolly
unique()

PinePlot <- 
  aes(x =
      y = ) + 
  #geom_point()+
  geom_jitter()+
  geom_smooth(
              # linetype = "dashed", # “blank”, “solid”, “dashed”, “dotted”, “dotdash”, “longdash”, “twodash” 
              # size = 15, 
              se = )

```

This plot shows how many bug bites you got when you tested different effectiveness of different bug sprays.

```{r Basic Plot, warning=FALSE}
BugSprayPlot <- |> ggplot()+
  aes(x = , y = , fill = ) +
  stat_boxplot()
```

In this figure, "eruptions" is the duration of the eruption, while "waiting" is the time to the next eruption after at the Old Faithful Geyser at Yellowstone National Park.

```{r Basic Plot, warning=FALSE}
OldFaithfulPlot <- ggplot() + 
  aes(x = , y = , color = ) + 
  geom_point()

```

Finally, this plot shows the miles per gallon for cars of different weights, with the gear coded to shape.

```{r Basic Plot, warning=FALSE}

```

## Relabeling Axes, Titles, Captions

First, we can make the text look nicer on our axes, titles, and add a caption.

```{r Change Labels}
PinePlotLabs <- PinePlot + 
  labs(y =  
       x = 
       title = 
       subtitle =
       caption = )

```

## Adding annotations/labels to the figure

What if we want to label on the figure, or label the points themselves?

```{r Annotate with a mapped variable using geom_text() and annotate() }
PinePlotLabs + 
  annotate() + 
  #geom_text(mapping = aes(label = Seed), position = "jitter")
  geom_label(mapping = )
  #ggrepel::geom_label_repel(aes(label = Seed))

```

## Visual Theme

The whole plot can have its appearance changed using themes.

```{r Set Whole-Plot Themes}
 + 
  #theme_bw()
#theme_classic()
#theme_light()
#theme_minimal()

```

## Changing the y and x-axis limits

Need more room for text, labels, additional geoms? Expand your axes using xlim() and ylim() or expand_limits()

```{r Change x and y axis limits}
 + 
  xlim() + 
  ylim() + 
  
```

## Change the scale of axes to log scale

Want to plot your log-transformed data but have the axis accurately represent the values, rather than the log of the values? Use the

```{r Log scale axis, and Change the tickmarks on a continuous x-axis}

# run before using
scale_x_ln <- function() # custom function for ln axis

PinePlotLabs + 
  # scale_x_log10() # native function for log10 axis
  scale_x_ln(breaks =)
```

## Relabel your axis ticks

```{r Relabel x-axis categories}
BugSprayPlot + 
  scale_x_discrete(name = ,
                      labels =)
```

## Changing Colors

### Setting colors manually

bit.ly/colors-r

https://r-charts.com/colors/

```{r Manual Colors}
customcolors <- c("cadetblue", "chartreuse2", "yellowgreen")

othercolors <- c("#F0FFFF", "#A52A2A", "#0000FF", "#FFB90F", "#556B2F", "#9932CC")


BugSprayPlot +
  #scale_fill_manual(values = customcolors)
  
```

### Setting colors using packages and palettes

```{r Viewing palette options}
library

RColorBrewer::

RColorBrewer::



```

Viridis color scales work for continuous or discrete variables, and look distinct when printed both in color and in black and white.

```{r Viridis}

library()

OldFaithfulPlot +
  
```

### Brewed palettes

```{r Color Palettes with brewer}
# have it select colors for you from a set
brewed_colors <- 

BugSprayPlot +
  

# Or! Just let it do it for you. 
BugSprayPlot +
  
```

https://github.com/EmilHvitfeldt/r-color-palettes

### Boutique color palettes ;)

```{r Wes Anderson Color Palettes}
library(wesanderson)


Wes_colors <- 

BugSprayPlot +
  scale_fill_manual(values = )

CarsPlot +
  aes(

```

https://github.com/jakelawlor/PNWColors

```{r Pacific Northwest Color Palettes}
library()
pnw_palettes

pnw_palette()

BugSprayPlot +
  scale_fill_manual(values = )
```

## Formatting Legends

```{r Formatting Legends}
BugSprayPlot + 
  theme() # Place legend at "bottom", "top", "left", or "right"
  
BugSprayPlot + 
   # Set legend type for each aesthetic: colorbar, legend, or none (no legend)

BugSprayPlot + 
  scale_fill_discrete(name = ,
                      labels = )
# Set legend title and labels with a scale function.

```

```{r}
glimpse()
CarsPlot + 
  aes( + 
  guides() # can turn off only one legend if you want
```

More: http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3/

## Combining multiple plots into one figure elegantly

```{r Combining multiple plots with patchwork}
library()

PinePlot +

patchwork <- (

patchwork & theme_bw()

```

<http://www.sthda.com/english/wiki/creating-and-saving-graphs-r-base-graphs>

Cheap way: Plots panel –\> Export –\> Save as Image or Save as PDF

## Exporting your plots

```{r Exporting Plots}

# set units to "in" or "cm", resolution (res) to 300 dpi for publication quality figures, set a size, and name your file.

tiff()
OldFaithfulPlot #Then, run all the code that goes with your figure! 
dev.off()  # Finally, the dev.off() line tells it you're all done and saves the figure. 


pdf() 

jpeg()
```

can do: pdf(“rplot.pdf”): pdf file png(“rplot.png”): png file jpeg(“rplot.jpg”): jpeg file postscript(“rplot.ps”): postscript file bmp(“rplot.bmp”): bmp file win.metafile(“rplot.wmf”): windows metafile

## Styling your Kable tables

```{r formatting kable tables}
MyLameTable <-
library()

kable(, 
      col.names = , # change your column names
      align = ,  # align either left "l", center "c" or right "r"
      caption = , 
      digits = , 
      # digits = c(2,5,1), # round up numbers without having to compute it yourself - can do differnet columns separately with c()
)
```

### Replace NA in your table with blanks or NA

```{r Replace NA with blanks or **}
opts <- options()
opts <-
```

```{r simple kableExtra}
library(kableExtra)
MyLameTable |>
  kbl() |>
  kable_classic()
```

```{r Using kableExtra styling}
kable(, booktabs = ) |>
  kable_styling(
    latex_options = ,
    # latex_options = c("striped", "scale_down"),
    font_size = ) |>
    row_spec() |> 
    column_spec() 
```

To export kable tables, the easiest way is to knit it to html, then copy and paste into a Word document.
