---
title: "Chi-Square and Tabular Tests"
format: html
editor: visual
Author: Amy King-Henry
lastmodified: Sept 10, 2026
---

# Tabular Experimental Designs in R

Chi-Square & goodness of fit are tests run on data with CATEGORICAL predictors AND CATEGORICAL responses. They are tests of association.

```{r}
load(file = "Demo Data Files/S4E2e.RData")
```

```{r}
#install.packages("rstatix")
library(tidyverse)
library(rstatix)

```

# Chi-squared analysis

- For when both IV and DV are categorical
- Tests for the "Goodness of fit" of a model
- Asks whether the OBSERVED patterns of occurrence differs from the EXPECTED patterns of occurrence (expected by chance)

**H0 - Data does not differ from expected distribution**

**HA - Data does differ significantly from expected distribution**

Data is formatted as a contingency table (a data matrix)

- Each row and column are a variable

- The data in each cell are counts of that combination of row and variable for example, if rows are colors, and columns are shapes, a cell might contain the count of red circles.

- NOTE: This is NOT "tidy" data! This is a different format and you should be able to transform from

### Example 1: Presence/Absence data of two heather species.

Do these two heather species occur *more frequently* together or *less frequently* together than expected by chance?

- Ecological importance:

  - More = evidence of mutualism or shared environmental conditions

  - Less = evidence of competition or opposing favored environmental conditions

```{r}
heather

#typeof(heather)
```

What data are in the rows?

- rows = whether *Erica cinerea* is present or absent

What data are in the columns?

- columns = whether *Calluna vulgaris* is present or absent.

What are the replicates/what do the data represent?

- Each count marks whether neither, one, or both are present.

How many replicate quadrats were surveyed?

```{r}
sum(heather)
```

137 quadrats were surveyed to see if one, both, or neither species were present.

How many quadrats had `Cal.vul` present?

```{r}
colSums(heather)
```

How many quadrats had `Eri.cin` present?

```{r}
rowSums(heather)
```

#### Chi-square test (for matrix)

The built-in R function for chi-square tests works just fine when you're dealing with a matrix.

```{r}
chisq.test(heather)
```

#### Transforming between matrix and data frame

Most of the data we work with however is in data frames! Let's transform our matrix to a data frame that keeps the essential information.

```{r}
heather_df <- tibble(heather) |> 
  mutate(Ercin = rownames(heather))

heather_df
```

We still need to pass only the values to the chi-square function, so we practice selecting only the

```{r}
# the bracket notation is an alternative way to select particular rows and columns - object[row, column]. The colon selects everything in the series

# heather_df[row, column]

heather_df[,1:2]
```

#### Chi-square test (data frame)

```{r}
TestResults <- chisq_test(heather_df[,1:2]) # the function can only take number data- so we have to input only the counts and exclude the Treatment column

TestResults
```

This gives us most of what we need! Remember that R automatically applies the Yates correction for 2x2 tables. What happens if we turn this off?

```{r}
# to turn off Yates, we do: 
chisq_test(heather_df[,1:2], correct = FALSE) 
```

#### Pearson Residuals

```{r}
PearResid <- pearson_residuals(TestResults)

PearResid
```

Remember, we can tell if any specific combination of rows/columns significantly differed from expected by examining the Pearson residuals. If any values are greater than 2 or less than -2, then we interpret that it was greater than 2 standard deviations from the expected counts, and therefore only a 5% likelihood that it occurred by chance.

#### Plots - Pearson Residuals

One way to view this is with a column plot. To make this plot, we need to convert it to tidy data.

```{r}
PlottablePearResid <- PearResid |> 
  as_tibble() |>  
  # PearResid comes out as a list - this converts it so that tidyverse can use it
  
  mutate(Treatment = c("Eri.cin.present", "Eri.cin.absent"))  |>  
  # add the treatment names back in manually to a new column - make sure they're in the same order as the input contingency table
  
  pivot_longer(1:2, names_to = "SpeciesPresAbs", values_to = "Residuals") # make it tidy to be easy to plot

PlottablePearResid
```

```{r}
ggplot(PlottablePearResid) + 
  aes(x = Treatment, y = Residuals, fill = SpeciesPresAbs) + 
  geom_col(position = "dodge") + 
  geom_hline(yintercept = c(-1.96,1.96), lty = "dotted") + 
  # add significance reference lines
  
  labs(y = "Pearson Residuals")
```

### Example 2: Bird and habitats

Adapt the method from the script above to run a chi-square test on whether birds associate with particular habitats.

```{r}
bird
```

```{r}
bird_df <- tibble(bird) |> 
  mutate(Species = rownames(bird))

bird_df
```

#### Chi-square test:

```{r}

```

#### Pearson Residuals/Plot:

```{r}

```

### Example 3: Invertebrate Habitats

Adapt the method from the script above to run a chi-square test on whether invertebrates associate with particular parts of a plant.

```{r}
inv.hab
```

#### Chi-square test:

```{r}

```

#### Pearson Residuals/Plot:

```{r}

```

## Goodness of Fit test

### Example 4: Peas and Goodness of Fit test

Sometimes you already have expected values, e.g., Mendelian genetics.

Peas come in smooth or wrinkled, green or yellow counts of each (based on dominant/recessive genotyping) are expected to be:

- 9 (green smooth),
- 3(green wrinkled),
- 3 (yellow smooth),
- 1 (yellow wrinkled)

If its different than expected, then it is possible that selection/differential survival is occurring.

```{r}
peas
```

```{r}
Obs.values <- peas
ratio <- c(9,3,3,1)
```

```{r}
my.good <- chisq.test(Obs.values, p = ratio, rescale.p = TRUE)
my.good
```
