---
title: "Great Graphs and Tables in ggplot and kable"
subtitle: "MCRS Quantitative Methods""
author: "Amy Henry"
date: "November 5, 2024"
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
install.packages("ggtext")
install.packages("PNWColors")
install.packages("viridis")
install.packages("wesanderson")

library(tidyverse)
library(ggtext)
```

So we all know by now how to make INFORMATIVE graphs using ggplot.

What if we want to make *beautiful* ones?

First let's make some basic plots to play with:

This is a plot of height of pine trees measured repeatedly at different ages.

```{r Basic Plot, warning=FALSE}
Loblolly
unique(Loblolly$Seed)

PinePlot <- Loblolly |> ggplot() +
  aes(x = age,
      y = height) + 
  #geom_point()+
  geom_jitter()+
  geom_smooth(color = "red", 
              # linetype = "dashed", # “blank”, “solid”, “dashed”, “dotted”, “dotdash”, “longdash”, “twodash” 
              # size = 15, 
              se = FALSE)
PinePlot
```

This plot shows how many bug bites you got when you tested different effectiveness of different bug sprays.

```{r Basic Plot, warning=FALSE}
BugSprayPlot <- InsectSprays |> ggplot()+
  aes(x = spray, y = count, fill = spray)+
  stat_boxplot()
BugSprayPlot
```

In this figure, "eruptions" is the duration of the eruption, while "waiting" is the time to the next eruption after at the Old Faithful Geyser at Yellowstone National Park.

```{r Basic Plot, warning=FALSE}
OldFaithfulPlot <- ggplot(faithful) + 
  aes(x = waiting, y = eruptions, color = waiting) + 
  geom_point()
OldFaithfulPlot
```

Finally, this plot shows the miles per gallon for cars of different weights, with the gear coded to shape.

```{r Basic Plot, warning=FALSE}
CarsPlot <- ggplot(mtcars) +
  aes(x = wt, y = mpg, shape = as.factor(gear)) + 
  geom_point() 
CarsPlot
```

## Relabeling Axes, Titles, Captions

First, we can make the text look nicer on our axes, titles, and add a caption.

```{r Change Labels}
PinePlotLabs <- PinePlot + 
  labs(y = "Height (m)", 
       x = "Tree Age (years)",
       title = "Loblolly Pine Growth by Age", 
       subtitle = "n = 14 trees",
       caption = expression(paste('Fourteen Loblolly pine trees (', italic("Pinus taeda"), ") were measured for height using standard field methods at five year intervals.")))

PinePlotLabs
```

## Adding annotations/labels to the figure

What if we want to label on the figure, or label the points themselves?

```{r Annotate with a mapped variable using geom_text() and annotate() }
PinePlotLabs + 
  annotate(geom = "text", x = 20, y = 25, label = "No trees were harmed in this study") + 
  #geom_text(mapping = aes(label = Seed), position = "jitter")
  geom_label(mapping = aes(label = Seed))
  #ggrepel::geom_label_repel(aes(label = Seed))

```

## Visual Theme

The whole plot can have its appearance changed using themes.

```{r Set Whole-Plot Themes}
PinePlotLabs + 
  #theme_bw()
#theme_classic()
#theme_light()
#theme_minimal()
theme_void()
```

## Changing the y and x-axis limits

Need more room for text, labels, additional geoms? Expand your axes using xlim() and ylim() or expand_limits()

```{r Change x and y axis limits}
PinePlotLabs + 
  xlim(0, 30) + 
  ylim(0, 80) + 
  expand_limits()
```

## Change the scale of axes to log scale

Want to plot your log-transformed data but have the axis accurately represent the values, rather than the log of the values? Use the

```{r Log scale axis, and Change the tickmarks on a continuous x-axis}

# run before using
scale_x_ln <- function(...) scale_x_continuous(..., trans = scales::log_trans()) # custom function for ln axis

PinePlotLabs + 
  # scale_x_log10() # native function for log10 axis
  scale_x_ln(breaks = c(1,5,10,15,20,25))
```

## Relabel your axis ticks

```{r Relabel x-axis categories}
BugSprayPlot + 
  scale_x_discrete(name = "Bug Spray Brands",
                      labels = c("Skin So Soft", "DEET", "Off!", "Picaridin", "Eucalyptus Oil", "Control"))
```

## Changing Colors

### Setting colors manually

bit.ly/colors-r

https://r-charts.com/colors/

```{r Manual Colors}
customcolors <- c("cadetblue", "chartreuse2", "yellowgreen", "wheat4", "violetred4", "turquoise")

othercolors <- c("#F0FFFF", "#A52A2A", "#0000FF", "#FFB90F", "#556B2F", "#9932CC")


BugSprayPlot +
  #scale_fill_manual(values = customcolors)
  scale_fill_manual(values = othercolors)
```

### Setting colors using packages and palettes

```{r Viewing palette options}
library(RColorBrewer)

RColorBrewer::display.brewer.all()

RColorBrewer::display.brewer.all(colorblindFriendly = T)



```

Viridis color scales work for continuous or discrete variables, and look distinct when printed both in color and in black and white.

```{r Viridis}

library(viridis)

OldFaithfulPlot +
  scale_color_viridis()
```

### Brewed palettes

```{r Color Palettes with brewer}
# have it select colors for you from a set
brewed_colors <- brewer.pal(6, "Set2")

BugSprayPlot +
  scale_fill_manual(values = brewed_colors)

# Or! Just let it do it for you. 
BugSprayPlot +
  scale_fill_brewer(palette = "Set2")
```

https://github.com/EmilHvitfeldt/r-color-palettes

### Boutique color palettes ;)

```{r Wes Anderson Color Palettes}
library(wesanderson)
wes_palettes

Wes_colors <- wes_palette("IsleofDogs1", 6, type ="discrete")

BugSprayPlot +
  scale_fill_manual(values = Wes_colors)

CarsPlot +
  aes(color = as.factor(cyl))+
  scale_color_manual(values = wes_palette("Moonrise1")) + 
  theme_dark()
CarsPlot
```

https://github.com/jakelawlor/PNWColors

```{r Pacific Northwest Color Palettes}
library(PNWColors)
pnw_palettes

pnw_palette("Starfish", 6, type = "discrete")

BugSprayPlot +
  scale_fill_manual(values = pnw_palette("Starfish", 6, type="discrete"))
```

## Formatting Legends

```{r Formatting Legends}
BugSprayPlot + 
  theme(legend.position = "bottom") # Place legend at "bottom", "top", "left", or "right"
  
BugSprayPlot + 
  guides(fill = "none") # Set legend type for each aesthetic: colorbar, legend, or none (no legend)

BugSprayPlot + 
  scale_fill_discrete(name = "New Legend Title",
                      labels = c("Skin So Soft", "DEET", "Off!", "Picaridin", "Eucalyptus Oil", "Control"))
# Set legend title and labels with a scale function.

```

```{r}
glimpse(mtcars)
CarsPlot + 
  aes(color = cyl) + 
  guides(color = "none", shape = "legend") # can turn off only one legend if you want
```

More: http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3/

## Combining multiple plots into one figure elegantly

```{r Combining multiple plots with patchwork}
library(patchwork)

PinePlot + OldFaithfulPlot / BugSprayPlot

patchwork <- (PinePlot | OldFaithfulPlot) / BugSprayPlot

patchwork & theme_bw()

```

<http://www.sthda.com/english/wiki/creating-and-saving-graphs-r-base-graphs>

Cheap way: Plots panel –\> Export –\> Save as Image or Save as PDF

## Exporting your plots

```{r Exporting Plots}

# set units to "in" or "cm", resolution (res) to 300 dpi for publication quality figures, set a size, and name your file.

tiff("MyFavoriteFigure.tif", width =4.5, height = 4, units = 'in', res = 300)
OldFaithfulPlot #Then, run all the code that goes with your figure! 
dev.off()  # Finally, the dev.off() line tells it you're all done and saves the figure. 


pdf("Fig2.pdf") 
print(BugSprayPlot)
dev.off() 


jpeg("Fig3.jpg", width = 350, height = "350")
print(PinePlotLabs)
dev.off()
```

can do: pdf(“rplot.pdf”): pdf file png(“rplot.png”): png file jpeg(“rplot.jpg”): jpeg file postscript(“rplot.ps”): postscript file bmp(“rplot.bmp”): bmp file win.metafile(“rplot.wmf”): windows metafile

## Styling your Kable tables

```{r formatting kable tables}
MyLameTable <- Loblolly |> group_by(age) |> summarise(mean = mean(height), sd = sd(height))
MyLameTable
library(knitr)

kable(MyLameTable, 
      col.names = c("Tree Age", "Mean Height", "Std.Dev."), # change your column names
      align = "c",  # align either left "l", center "c" or right "r"
      caption = "Tables can have captions too!", 
      digits = 2, 
      # digits = c(2,5,1), # round up numbers without having to compute it yourself - can do differnet columns separately with c()
)
```

### Replace NA in your table with blanks or NA

```{r Replace NA with blanks or **}
opts <- options(knitr.kable.NA = "")
opts <- options(knitr.kable.NA = "**")
```

```{r simple kableExtra}
library(kableExtra)
MyLameTable |>
  kbl(caption = "Mean Tree Height at Ages") |>
  kable_classic(full_width = F, html_font = "Cambria")
```

```{r Using kableExtra styling}
kable(MyLameTable, booktabs = TRUE) |>
  kable_styling(
    latex_options = "striped",
    # latex_options = c("striped", "scale_down"),
    font_size = 8) |>
    row_spec(2, bold = TRUE, italic = TRUE) |> 
    column_spec(3,background = "gray") 
```

To export kable tables, the easiest way is to knit it to html, then copy and paste into a Word document.
