---
title: "Logistic Regression"
author: Amy King-Henry
datemodified: Sept 2, 2026
format: html
editor: visual
---

# Logistic Regression

Logistic Regression is for when your **response variable is categorical, but your predictor is continuous,** and mostly when it can only be in one of two forms, aka, **binomial**. This type of test is concerned with the **ODDS** of getting one state or the other of your binomial variable, depending on the state of a quantitative predictor variable.

```{r}
load(file = "Demo Data Files/S4E2e.RData")

guppy <- read.csv(file = "Demo Data Files/chap17f9_1GuppyColdDeath.csv")
lions <- read.csv(file = "Demo Data Files/chap17e1LionNoses.csv")
earthworms <- read.csv(file = "Demo Data Files/chap17q17EarthwormsAndNitrogen.csv")
```

Load packages

```{r}
library(tidyverse)
```

## Simple Logistic Regression

### Example 1: Newts and plant cover

What are the chances of finding a newt depending on the level of plant cover?

#### Step 1: tidy your data

This dataset contains survey information on dozens of different ponds, and then whether newts were found in them or not (presence)

The "macro" column indicates the percent cover of vegetation on the pond.

```{r}
glimpse(gcn)
```

#### Step 2: Visualize the data

If we try just a scatter, we're going to miss out on observations because they all overlap...

```{r}
ggplot(gcn)+
  aes(macro, presence)+
  geom_point()
```

It's better if we add some jitter so we can see all the observations.

```{r}
ggplot(gcn)+
  aes(x = macro, y = presence)+
  geom_jitter(height = 0.1, shape = 1)
```

Another way to do it is with geom_count(), which sizes the circle based on how many overlapping observations it contains.

```{r}
newt_plot <- ggplot(gcn)+
aes(macro, presence)+
  geom_count()

newt_plot
```

What variables are here?

1.  macro = macrophyte cover in the pond
2.  presence = at least one newt was found in that pond

#### Step 3: Run the logistic regression model

`glm()` is the function that will run generalized linear models, which includes logistic regression and its binary response variable.

The setup of this function is the same as for `lm(),` except that we specify a **family**.

**Family** tells what kind of distribution R should expect - in this case, we want binomial because that is how our response variable is distributed

**Remember the components of the GLM are:**

1.  The [random component]{.underline} (the distribution of the data) - here it's binomial.

2.  The [systematic component]{.underline} (the formula that represents our hypothesis for the relationship between the variables).

3.  The [link function]{.underline} (the function that transforms the relationship to make it linear)

```{r}
newt.glm <- glm(presence ~ macro, data = gcn, family = binomial(link = "logit"))

summary(newt.glm)
```

Is macro statistically significant? What does that mean?

#### Step 4: Calculate the D^2^ value

Instead of R^2^ we look at D^2^ for Logistic Regression, which we can interpret similarly - the proportion of variance explained by the predictor variables.

```{r}

newt.glm$deviance
newt.glm$null.deviance


1 - (newt.glm$dev /newt.glm$null)
```

Not that good of a predictor! Let's look at the plot:

```{r}
newt_plot+
  stat_smooth(method="glm", method.args=list(family="binomial"), se=FALSE)
```

What is the likelihood of finding a newt if you have 50% cover of macrophytes on your pond?

```{r}
#this is just our model from above again
newt.glm

nd <- data.frame(macro = c(40, 50, 60)) # This creates a new data frame with some x values

nd # let us look at them


predict(newt.glm, newdata = nd, type = "response") # This function uses the model and the new data (that we generated a second ago) to make a prediction for the values. 
```

### Example 2: Guppy Cold Death

data from : Pitkow, R. B. 1060. Biological Bulletin 119: 231-245.

We saw this example in our lecture.

This dataframe is formatted weird, so I've taken care of the cleaning here.

```{r}
glimpse(guppy)

guppy <- guppy %>% 
  mutate(exposureDurationMin = as.numeric(as.character(exposureDurationMin)), 
         mortality = as.numeric(as.character(mortality)))

glimpse(guppy)
```

#### Step 1: Plot the Data

```{r}
guppyplot <- 

guppyplot
```

#### Step 2: make the model. Is duration exposure a significant predictor of guppy death?

```{r}
guppymod <- glm()

summary(guppymod)

```

#### Step 3: How much of the variation in guppy death is explained by how long they were exposed to freezing temperatures?

```{r}
# calculate the D^2

```

#### Step 4: Plot your data again and add a smoothed logistic regression line.

```{r}
guppyplot + 
  stat_smooth(method="glm", method.args=list(family="binomial"), se=FALSE)


```

#### Step 5: What is the LD50 (or lethal dose/duration) at with 50% of the guppies die?

```{r}

```

## Multiple Logistic Regression - using STEPWISE regression

Back to our newt model, that really was a terrible fit though! Let's see if we can get a better model using the `add1()` function for stepwise multiple logistic regression.

Stepwise regression helps choose which variables to include in a model by adding them in one at a time and testing how much better they make the fit.

We start with a null model - our formula is just `response ~ 1` , with the number 1 as a stand-in for "nothing at all".

We then use the `add1()` function on that model, give it the dataset and a test (`Chisq` here), and what it returns is the results of each of the variables in that dataset if they were added to the model.

e.g.,

- `presence ~ area`

- `presence ~ dry`

- `presence ~ water` etc.

We can then compare the fit of those models using AIC (Akaike Information Criterion) and the reported p-values from our chi-square test.

If the chi-square value is \<0.05 and the AIC score is LOWER than the first model (here its the <none> model) by more than 2, then its a significant predictor.

```{r}
#null model
newt_mod <- glm(presence ~ 1, data = gcn, family = binomial)

# this function tests all the other variables for their
add1(newt_mod, gcn, test = "Chisq")
```

We look to see which one has the GREATEST reduction in the AIC score, then add that variable to the systematic component of our model in place of the 1 we had before.

This new model is now a one variable logistic regression. But why stop there? What if a second factor is explanatory?

We can run our add1() function again now that we've updated our model to see what model fits look like with a second variable added.

E.g.,

- `presence ~ HSI + area`

- `presence ~ HSI + dry`

- `presence ~ HSI + water` etc.

And we do our comparison of AIC scores again.

```{r}
newt_mod <- glm(presence ~ HSI, data = gcn, family = binomial)
add1(newt_mod, gcn, test = "Chisq")
```

repeat...

```{r}
newt_mod <- glm(presence ~ HSI + area, data = gcn, family = binomial)
add1(newt_mod, gcn, test = "Chisq")
```

```{r}
newt_mod <- glm(presence ~ HSI + area + land, data = gcn, family = binomial)
add1(newt_mod, gcn, test = "Chisq")
```

```{r}
newt_mod <- glm(presence ~ HSI + area + land + dry, data = gcn, family = binomial)
add1(newt_mod, gcn, test = "Chisq")
```

```{r}
newt_mod <- glm(presence ~ HSI + area + land + dry + bird, data = gcn, family = binomial)
add1(newt_mod, gcn, test = "Chisq")
```

Let's check our our final model!

```{r}
summary(newt_mod)
```

And calculate our D\^2

```{r}
1 - (newt_mod$dev /newt_mod$null)
```

better than before!

## 
